---
title: "Examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biogate)
```

Practical recipes. Each one is a short, real task. For the full list of sources
and a valid example id for each, see the
[sources cookbook](sources.html).

## Clean a column of identifiers

Split a vector into the ones that pass and the ones that do not, keeping order.

```{r}
ids <- c("MONDO:0005148", "mondo:5148", "MONDO:0018076", "banana")
ok <- is_valid_id(ids, source_db = "mondo")

ids[ok] # valid
ids[!ok] # invalid
```

## Repair what can be repaired

`check_id()` returns a `suggestion` for an input that is invalid but mappable,
such as a lowercase prefix. Use it to fix a column instead of dropping it.

```{r}
res <- check_id(ids, source_db = "mondo")
res$fixed <- ifelse(res$valid, res$normalized, res$suggestion)
res[c("input", "valid", "fixed")]
```

`banana` has no repair, so its `fixed` value is `NA`.

## Existence, not just shape

`pattern` mode only checks the shape. To check that an id actually exists, use a
pinned snapshot (`cache`) or the live source (`remote`).

```{r}
# offline, against the sample snapshot that ships with the package
is_valid_id("MONDO:9999999", source_db = "mondo", how = "cache", version = "sample")
```

`MONDO:9999999` is well-formed, so it passes `pattern`, but it is not in the
snapshot. The live check needs a network, so it is shown but not run here:

```{r, eval = FALSE}
is_valid_id("ENSG00000139618", source_db = "ensembl", how = "remote")
```

A failed `remote` lookup raises an error. It never returns a silent `FALSE`, so
a network problem cannot be mistaken for an absent id.

## Species context

For species-aware sources, pass `species`. For Ensembl the species is encoded in
the id, so even `pattern` mode can reject a well-formed id from the wrong
species.

```{r}
is_valid_id("ENSMUSG00000059552", source_db = "ensembl", species = "mus_musculus")
is_valid_id("ENSMUSG00000059552", source_db = "ensembl", species = "homo_sapiens")
```

## Guard a pipeline with checkmate

The checkmate-style adapters check, assert, or test a vector against a source.

```{r}
# a friendly message instead of TRUE when something is off
check_valid_id(c("MONDO:0005148", "mondo:5148"), "mondo")

# a single logical, handy in `if` and `stopifnot`
test_valid_id("MONDO:0005148", "mondo")
```

## Validate a data frame with assertr or validate

`id_predicate()` returns an elementwise predicate, which is what data-frame
validators expect.

```{r}
is_mondo <- id_predicate("mondo")
ids[is_mondo(ids)]
```

Drop it straight into a pipeline. These need the `assertr` or `validate`
package, so they are shown but not run here:

```{r, eval = FALSE}
# assertr, inside a dplyr pipeline
df |> assertr::assert(id_predicate("mondo"), term)

# validate
rules <- validate::validator(good_terms = id_predicate("mondo")(term))
validate::confront(df, rules)
```

## Validate a Shiny input

`sv_biogate()` returns a shinyvalidate rule. It returns `NULL` for a valid input
and a message otherwise.

```{r}
rule <- sv_biogate("mondo")
rule("MONDO:0005148") # NULL, valid
rule("mondo:5148") # a message
```

## Discover sources in code

You never need to hard-code a key or guess an example. `source_info()` lists
every source with a valid example and its supported modes.

```{r}
source_info()
```
